### 2013-05-06 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding jacobi preconditioned fuel assembly data to doc

%%---------------------------------------------------------------------------%%
\section{The Neutron Transport Equation}
\label{sec:transport_eq}
As a starting point we define the time-independent neutron transport
equation \citep{lewis_computational_1993}:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
  q(\vec{r},\hat{\Omega},E)\:,
  \label{eq:general_transport}
\end{multline}
with the variables defined as:
\begin{itemize}
\item $\vec{r}$ - neutron spatial position
\item $\hat{\Omega}$ - neutron streaming direction with radial
  component $\mu$ and azimuthal component $\omega$
\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
  scattering
\item $E$ - neutron energy
\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
\item $\sigma(\vec{r},E)$ - total interaction cross section
\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
  of scattering from direction $\hat{\Omega}'$ into an angular domain
  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
  $E'$ to an energy domain $dE'$ about energy $E$
\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
\end{itemize}
For this work, it is sufficient to formulate
Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
  \frac{q(x,E)}{4 \pi}\:,
  \label{eq:cart_1d_transport}
\end{multline}
where the angular component of the solution is no longer dependent on
the azimuthal direction of travel and an isotropic source of neutrons
is assumed. In addition, for fission systems the eigenvalue form of the transport
equation is:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
  \label{eq:eigenvalue_transport}
\end{multline}
with the addition variables defined as
\begin{itemize}
\item $k$ - multiplication factor
\item $\chi(E)$ - fission neutron energy spectrum
\item $\nu$ - average number of neutrons per fission
\item $\sigma_f(r,E')$ - fission cross section\:.
\end{itemize}
In 1-dimensional cartesian geometry,
Eq~(\ref{eq:eigenvalue_transport}) becomes:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
  dE' + \frac{q(x,E)}{4 \pi}\:.
  \label{eq:cart_1d_eigenvalue}
\end{multline}

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Before we can move on to solving the $SP_N$ equations with both
conventional and Monte Carlo methods, we must first understand their
spectral character to ensure the MCSA will be applicable. As presented
in Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
iteration matrix must be less than unity to ensure
convergence. Therefore, we will compute the spectral radius of the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $4 \times 4 \times 4$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}
Mesh element sizes were the same in all cardinal directions with
reflecting conditions on each boundary. For the energy parameter, 1
and 10 energy groups were used with upscatter/downscatter varied for
the 10 group case. For the downscatter cases, all groups could scatter
to all lower energy groups with the cross sections given in
Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
groups could scatter to all other energy groups with different cross
sections for upscatter and downscatter used. A uniform isotropic
source of neutrons is given in each group as well.

@@

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10uspj}
\end{table}
It is readily apparent from the tabulated data the point Jacobi
preconditioning is insufficient as a preconditioning mechanism for
problems of order $SP_5$ or higher. In addition, as upscatter is
introduced into the system, a much larger spectral radius is observed
at lower $SP_N$ orders, signaling a need for a better preconditioning
strategy to both ensure and improve covergence for Monte Carlo
methods.

@@

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a $17 \times 17$ hot zero-power pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical access
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_radial_mat}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Each fuel pin is resolved by a $2 \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.25in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.}  \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and core
      structures on the top and bottom of the assembly.}}
  \label{fig:problem3_axial_mat}
\end{figure}
Significant geometric details are contained in the model including
spacer grids, fuel pins with homogenized cladding and gas gap, and
core plenum. Table~\ref{tab:problem3_parameters} gives the primary
design parameters for the fuel assembly calculations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Parameter}} & 
      \multicolumn{1}{l}{\textbf{Value}} \\
      Power Level & 0 MW \\
      Inlet Temperature & 326.85C \\
      Fuel Temperature & 600C \\
      Boron Concentration & 1300 ppm \\
      Moderator Density & 0.743 g/cc \\
      Helium Density & \sn{1.79}{-4} g/cc \\
      Zirconium Density & 6.56 g/cc \\
      Stainless Steel Density & 8.0 g/cc \\
      Inconel Density & 8.19 g/cc \\
      UO2 Density & 10.257 g/cc \\
      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
      assembly criticality calculation.}}
  \label{tab:problem3_parameters}
\end{table}

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\ve{\phi_0} =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\ve{M} \ve{\phi}^{n+1} = \frac{1}{k^n} \ve{F} \ve{\phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \ve{F} \ve{\phi}^{n+1} d\ve{r}}{\int
      \ve{F} \ve{\phi}^n d\ve{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\ve{M}$ is the transport operator generated on the left-hand
side of the $SP_N$ discretization and $\ve{F}$ is the fission matrix.
This problem is significantly more complicated that the simple test
problem used for the previous spectral analysis. Fission has been
introduced into the set of equations and the addition of moderator
into the system will increase the amount of scattering, creating a
significantly more difficult problem.

\subsection{Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}

\subsection{MCSA Break Down}
\label{subsec:mcsa_break_down}

@@

>>>>

%%---------------------------------------------------------------------------%%
\section{The Neutron Transport Equation}
\label{sec:transport_eq}
As a starting point we define the time-independent neutron transport
equation \citep{lewis_computational_1993}:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
  q(\vec{r},\hat{\Omega},E)\:,
  \label{eq:general_transport}
\end{multline}
with the variables defined as:
\begin{itemize}
\item $\vec{r}$ - neutron spatial position
\item $\hat{\Omega}$ - neutron streaming direction with radial
  component $\mu$ and azimuthal component $\omega$
\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
  scattering
\item $E$ - neutron energy
\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
\item $\sigma(\vec{r},E)$ - total interaction cross section
\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
  of scattering from direction $\hat{\Omega}'$ into an angular domain
  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
  $E'$ to an energy domain $dE'$ about energy $E$
\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
\end{itemize}
For this work, it is sufficient to formulate
Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
  \frac{q(x,E)}{4 \pi}\:,
  \label{eq:cart_1d_transport}
\end{multline}
where the angular component of the solution is no longer dependent on
the azimuthal direction of travel and an isotropic source of neutrons
is assumed. In addition, for fission systems the eigenvalue form of the transport
equation is:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
  \label{eq:eigenvalue_transport}
\end{multline}
with the additional variables defined as
\begin{itemize}
\item $k$ - multiplication factor
\item $\chi(E)$ - fission neutron energy spectrum
\item $\nu$ - average number of neutrons per fission
\item $\sigma_f(r,E')$ - fission cross section\:.
\end{itemize}
In 1-dimensional cartesian geometry,
Eq~(\ref{eq:eigenvalue_transport}) becomes:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
  dE' + \frac{q(x,E)}{4 \pi}\:.
  \label{eq:cart_1d_eigenvalue}
\end{multline}

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Before we can move on to solving the $SP_N$ equations with both
conventional and Monte Carlo methods, we must first understand their
spectral character to ensure that MCSA will be applicable. As presented
in Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
iteration matrix must be less than unity to ensure
convergence. Therefore, we will compute the spectral radius of the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $4 \times 4 \times 4$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}
Mesh element sizes were the same in all cardinal directions with
reflecting conditions on each boundary. For the energy parameter, 1
and 10 energy groups were used with upscatter/downscatter varied for
the 10 group case. For the downscatter cases, all groups could scatter
to all lower energy groups with the cross sections given in
Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
groups could scatter to all other energy groups with different cross
sections for upscatter and downscatter used. A uniform isotropic
source of neutrons is given in each group as well.

@@

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10uspj}
\end{table}
It is readily apparent from the tabulated data that point Jacobi
preconditioning is insufficient. For problems of order $SP_5$ and
$SP_7$, the method will not converge at all and for the upscatter case
with $SP_3$ the spectral radius is still quite large for large $P_N$
orders. This data signals a need for a better preconditioning strategy
to both ensure and improve covergence for Monte Carlo methods.

@@

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a $17 \times 17$ hot zero-power pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical access
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Each fuel pin is resolved by a $2 \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.25in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.}  \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and core
      structures on the top and bottom of the assembly.}}
  \label{fig:problem3_axial_mat}
\end{figure}
Significant geometric details are contained in the model including
spacer grids, fuel pins with homogenized cladding and gas gap, and
core plenum. Table~\ref{tab:problem3_parameters} gives the primary
design parameters for the fuel assembly calculations. Group cross
sections and other discrete nuclear data are generated as needed by a
cross-section processing module.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Parameter}} & 
      \multicolumn{1}{l}{\textbf{Value}} \\
      Power Level & 0 MW \\
      Inlet Temperature & 326.85C \\
      Fuel Temperature & 600C \\
      Boron Concentration & 1300 ppm \\
      Moderator Density & 0.743 g/cc \\
      Helium Density & \sn{1.79}{-4} g/cc \\
      Zirconium Density & 6.56 g/cc \\
      Stainless Steel Density & 8.0 g/cc \\
      Inconel Density & 8.19 g/cc \\
      UO2 Density & 10.257 g/cc \\
      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
      assembly criticality calculation.}}
  \label{tab:problem3_parameters}
\end{table}

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate converge and remain unchanged throughout the
computation while the addition of fission to the system will only
modify the source of neutrons and not affect Monte Carlo transport.

\subsection{Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Even if the problem may be
ill-conditioned, we do expect convergence of MCSA. To investigate
further, a block Jacobi preconditioned Richardson iteration was used
to solve the same problem. Figure~\ref{fig:block_jacobi_res_richardson}
gives the residual infinity norm as a function of iteration for the
Richardson linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.}  \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration. However,
convergence is achieved meaning that the criteria needed to satisfy
MCSA convergence has also been met. The effective slope of the
convergence curve in Figure~\ref{fig:block_jacobi_res_richardson} will
give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). Based on these results, it then
appears that even if the simple criteria of a spectral radius of less
than one is met and the Richardson iteration will converge with the
same preconditioning, MCSA still may not converge. We then expect the
issue to reside with the Neumann-Ulam solve providing the correction
as the Richardson iteration is known to provide the correct
result. Furthermore, the fact that the spectral radius is less than
one means that the stochastic histories in the block Jacobi
preconditioned Neumann-Ulam method are eventually being terminated by
the weight cutoff as no artificial absorption was used. Based on this,
the correction being generated by the Neumann-Ulam solve must be
incorrect.

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi calculations
used a one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use a simpler homogoneous 2-dimensional
one-speed neutron diffusion system to isolate this behavior. In this
system, we can vary the cross sections in order to achieve varying
spectral radii.

@@

\subsubsection{MCSA Relaxation Parameters}
\label{subsubsec:spn_mcsa_relaxation}
